Disparities in Climate-Induced Health Outcomes in the Greater Toronto Area.

### 2022-08-08 outputs/paper/paper.qmd
Updates to paper

As climate change worsens, the danger it presents to human health becomes more and more dangerous [@citenasaclimatewarn]. The health impacts due to climate change are already being demonstrated in parts of Canada. Canada is warming at almost twice the global rate and if this continues, emissions will exceed 2 degrees Celsius worldwide [@citehealthcanada].

@@

Urgent action is needed and the science shows that it will make a tangible difference \[\@\]. The health impacts from climate change can be mitigated by improving efforts both in societies and health systems in Canada \[\@\]. Changes in measures can prevent health impacts and build health systems and facilities that are more resilient to climate change \[\@\].

@@

The data for temperature was processed using R studio and R [@citeR]. The 2002 to 2022 temperature data was combined together. The data set initially contained 31 variables. From these the city, date, year, month, day, maximum temperature, minimum temperature, and mean temperature were selected using the dplyr package [@citedplyr]. Then any NA values from the maximum, minimum, and mean temperature columns were dropped.

@@

Table 1 shows some of the rows of the Temperature data from 2002 to 2022.

@@

Table 2 shows some of the rows of the Pollution data from 2002 to 2020. Variable "Mean Pollution" represents the mean pollution levels of the following pollutant on the following day of the month and year.

@@

temp_2002_2022 %>% filter((month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar") & (year >= 2002 & year <= 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

temp_2002_2022 %>% filter((month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | 
                             month_name == "Mar") & (year > 2012 & year <= 2022)) %>% ggplot(aes(x=year, y=max_temp, 
                                                                                                  color=factor(month_name, level = level_order))) + 
  geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + 
  theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

temp_2002_2022 %>% filter((month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") & (year >= 2002 & year <= 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

temp_2002_2022 %>% filter((month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") & (year > 2012 & year <= 2022)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

temp_2002_2022 %>% filter((month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") & (year >= 2002 & year <= 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

temp_2002_2022 %>% filter((month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") & (year > 2012 & year <= 2022)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) + scale_x_discrete(breaks = 0:2100)

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "gray") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light blue") +
  ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light pink") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#9AD67F") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#847FD6") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

>>>>

As climate change worsens, the danger it presents to human health becomes more dangerous [@citenasaclimatewarn]. The health impacts due to climate change are already being demonstrated in parts of Canada. Canada is warming at almost twice the global rate and if this continues, emissions will exceed 2 degrees Celsius worldwide [@citehealthcanada].

@@

Urgent action is needed and the science shows that it will make a tangible difference \[\@citeipcc\]. The health impacts from climate change can be mitigated by improving efforts both in societies and health systems in Canada \[\@citeipcc\]. Changes in measures can prevent health impacts and build health systems and facilities that are more resilient to climate change \[\@citeipcc\].

@@

The data for temperature was processed using R studio and R [@citeR]. The 2002 to 2022 temperature data was combined together. The data set initially contained 31 variables. From these the city, date, year, month, day, maximum temperature, minimum temperature, and mean temperature were selected using the dplyr package [@citedplyr].

@@

@@

Table 1 shows some rows of the Temperature data from 2002 to 2022. Table 2 shows some of the rows of the Pollution data from 2002 to 2020. Variable "Mean Pollution" represents the mean pollution levels of the following pollutant on the following day of the month and year.

@@

temp_2002_2022 %>% filter((month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar") & (year >= 2002 & year < 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) 
```

@@

temp_2002_2022 %>% filter((month_name == "Dec" | month_name == "Jan"| month_name == "Feb" | month_name == "Mar") & (year >= 2012 & year < 2022)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name)

@@

temp_2002_2022 %>% filter((month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") & (year >= 2002 & year < 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) 
```

@@

temp_2002_2022 %>% filter((month_name == "Apr" | month_name == "May" | month_name == "Jun" | month_name == "Jul") & (year >= 2012 & year < 2022)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name) 
```

@@

temp_2002_2022 %>% filter((month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") & (year >= 2002 & year < 2012)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name)
```

@@

temp_2002_2022 %>% filter((month_name == "Aug" | month_name == "Sep" | month_name == "Oct" | month_name == "Nov") & (year >= 2012 & year < 2022)) %>% ggplot(aes(x=year, y=max_temp, color=factor(month_name, level = level_order))) + geom_point(alpha = 0.4) + labs(color="Month") + xlab("Year") + ylab("Maximum Temperature") + theme_bw() + facet_grid(. ~ month_name)
```

@@

pollution %>% filter(pollutant == "NO" & is.na(month_name) == "FALSE") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "gray") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2" & is.na(month_name) == "FALSE") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light blue") +
  ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3" & is.na(month_name) == "FALSE") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light pink") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO" & is.na(month_name) == "FALSE") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#9AD67F") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2" & is.na(month_name) == "FALSE") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#847FD6") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```
